Efficient Compilation of High Level Python Numerical Programs with Pythran

### 2014-09-02 doc/papers/cgo2014/paper.tex
CGO: Rephrase effects analysis

To inter procedurally compute the read-write effects, intra procedural effects
are first computed. Then we create a directed graph $G$ with as many nodes as
functions. Each node contains a list of integer representing the indices of the
formal argument that are modified intra-procedurally. Then we add an edge
between two nodes, labeled by an integer tuple $(i, j)$ if $i^\text{th}$ formal
argument of the source is passed as $j^\text{th}$ effective argument of the
sink. The goal is to prune all edges from $G$. To do so, given a node $n$ that
writes $i$, any outgoing edge $(i, j)$ can be safely removed, as we know for
sure that the $i^\text{th}$ parameter is written. Similarly, an incoming edge
$(j, i)$ can be removed and $j$ is added to the source's list. Iteratively
applying this process removes all edges that can propagate a write, the only
remaining edges propagate a read and can be ultimately removed.

Let us illustrate the algorithm on a simple case. Python code source and the
initial effect graphs are illustrated on Figure~\ref{fig:argument-effects}. The
only node with write effects is \texttt{bar} that writes \texttt{n} through a
subscript. Outgoing edges (1) and (2) can be removed because they are
outgoing edges of \texttt{bar} starting by \texttt{n}. Incoming edge (0) ends
by \texttt{n} that is written by \texttt{bar} so it is possible to prune it
while adding the start \texttt{n} to node \texttt{root}. No edge left, and we
know that both \texttt{root} update \texttt{r} and \texttt{bar} update \texttt{n},
but \texttt{foo} doesn't update its argument.

>>>>

To interprocedurally compute the read-write effects, intra procedural effects
are first computed. Then we create a directed graph $G$ with as many nodes as
functions, and with as many edges as formal parameters between two nodes. Local
variables used as arguments do not trigger an edge.
Edges are labeled by an integer tuple $(i, j)$ if $i^\text{th}$ formal
argument of the source is passed as $j^\text{th}$ effective argument of the
sink. The analysis is iteratively pruning all edges from $G$ by considering any
node for which the intraprocedural analysis shown that an argument $i$ is 
written. Since the parameter is modified, any outgoing edge $(i, j)$ for this 
node will not add any information to the analysis and then can be safely 
removed. Incoming edges $(j, i)$ are used to propagate the effects along the 
call graph: $j$ is added to the source's list of argument written and the source
node is added to the list of nodes waiting to be processed. Iteratively
applying this process removes all edges that can propagate a write, when no edge
can be removed, the remaining ones propagate a read and can be ultimately 
removed.

Let us illustrate the algorithm on a simple case. Python code source and the
initial effect graphs are illustrated on Figure~\ref{fig:argument-effects}. The
only node with write effects is \texttt{bar} that modifies \texttt{n} through a
subscript. Outgoing edges (1) and (2) can be removed because they are
outgoing edges of \texttt{bar} starting by \texttt{n}. Incoming edge (0) ends
by \texttt{n} that is written by \texttt{bar}, so it propagates the write 
effects to the source node corresponding to the function \texttt{root}. After
removing this edge, the only remaining edges are propagating read and the 
analysis ands. The analysis informs that both \texttt{root} update \texttt{r} 
and \texttt{bar} update \texttt{n}, but \texttt{foo} doesn't update its 
argument.
